---
title: "PA1_template"
author: "Gene Livingston"
date: "Tuesday, May 03, 2016"
output: html_document
---
# Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

# Source Data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

Dataset: <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip>


## The variables included in the dataset are:

* steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* date: The date on which the measurement was taken in YYYY-MM-DD format
* interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

# Loading and preprocessing the data

Load the data and perform pre-processing in order to remove NA values. Also reformat the date data.

```{r}


activity <- read.csv("activity.csv")
activity$day <- weekdays(as.Date(activity$date))
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
clean <- activity[!is.na(activity$steps),]
```

# What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
2. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum, )
colnames(sumTable)<- c("Date", "Steps")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

# Calculate and report the mean and median of the total number of steps taken per day

```{r}
mea <- as.integer(mean(sumTable$Steps))
med <- as.integer(median(sumTable$Steps))
mea
med
```

**The average number of daily steps was `r mea` steps.**

**The median number of daily steps was `r med` steps.**


# What is the average daily activity pattern?

## Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
library(plyr)
library(ggplot2)
tidy <- activity[!is.na(activity$steps),]

intervalTable <- ddply(tidy, .(interval), summarize, Avg = mean(steps))

p <- ggplot(intervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```


# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxSteps <- max(intervalTable$Avg)
maxs <- as.integer(maxSteps)
itab <- intervalTable[intervalTable$Avg==maxSteps,1]
itab

```

**The maximum number of steps (5-minute interval) was `r maxs` steps.**

**The interval with the maximum number of steps was the `r itab` interval.**


# Imputing missing values

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
nrow(activity[is.na(activity$steps),])
```

**The total number of rows with steps = 'NA' is 2304.**

## Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

**Replace the missing steps with the average interval (based on the average for that day).**

```{r}
avgTable <- ddply(clean, .(interval, day), summarize, Avg = mean(steps))

nadata<- activity[is.na(activity$steps),]
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```

# Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

mergeData <- rbind(clean, newdata2)
```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum, )
colnames(sumTable2)<- c("Date", "Steps")

newMean <- as.integer(mean(sumTable2$Steps))
newMean
```

```{r}
newMed <- as.integer(median(sumTable2$Steps))
newMed
```

```{r}
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Black")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("black", "grey") )
```


**The new mean of the imputed data is `r newMean ` steps compared to the old mean of `r mea` steps. That creates a difference of `r newMean - mea` steps on average per day.**

**The new median of the imputed data is `r newMed` steps compared to the old median of `r med` steps. That creates a difference of `r newMed - med` steps for the median.**  

**The shape of the distribution has not changed.**


# Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).


```{r}
library(lattice) 
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))

xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```


**Yes, the step activity trends are different based on whether the day occurs on a weekend or not. Likely, there is more time to exercise on the weekends.**










